OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2018-01-16 manuscript_nar.tex
Minor edits finalizing revision R2.

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~Kbp.
\textit{OneD} is implemented as an R package available at
http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720, see Supplementary Table~1), that were performed
in either B cells or ES cells, and from either CRG or UCSD. The simulation
strategy was the following: First we selected uniformly at random the
amount of copy number break points (from 3 to 10). We then placed the
break points along chromosomes 18 and 19 uniformly at random. Next, we
assigned a copy number (2, 3, 4 or 10 with equal probability) to each
segment delimited by the break points. We computed the outer product of
this simulated copy number profile and multiplied it element-wise with the
original contact matrices of chromosomes 18 and 19. The resulting matrices
were used as input for correction methods. For each simulation (100 in
total), we measured the pairwise reproducibility score defined by
\cite{yan2017hicspector} and computed the average for pairs from the same
cell type minus the average for pairs from different cell types.

@@

The originality of \textit{OneD} lies in the estimation from a single
variable of the explicit biases with a Negative Binomial model and of the
copy number. This allows greater running speed, while preserving a good
performance on samples with karyotypic aberrations. One of the reasons why
\textit{OneD} is able to better highlight the biological distinctions
between samples is that it only corrects the copy number if specifically
requested by the user. The impact of copy number variations on
normalization is rather opaque, especially if they are treated as implicit
biases (Figure~\ref{fig:cnv_correction}B). Treating them as explicit
biases with optional removal seems to be an overall safer strategy. In
homogeneous data sets, correcting for the copy number can blur the
biological signal, but when karyotypes are variable or unstable, it may
increase the reproducibility (Figure~\ref{fig:cnv_correction}A).

>>>>

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~Kbp.
The \textit{OneD} R package and the data used here are
available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720, see Supplementary Table~1), that were performed
in either B cells or ES cells, and from either CRG (Center for Genomic
Regulation) or UCSD (University of California San Diego). The simulation
strategy was the following: First we selected uniformly at random the
amount of copy number break points (from 3 to 10). We then placed the
break points along chromosomes 18 and 19 uniformly at random. Next, we
assigned a copy number (2, 3, 4 or 10 with equal probability) to each
segment delimited by the break points. We computed the outer product of
this simulated copy number profile and multiplied it element-wise with the
original contact matrices of chromosomes 18 and 19. The resulting matrices
were used as input for correction methods. For each simulation (100 in
total), we measured the pairwise reproducibility score defined by
\cite{yan2017hicspector} and computed the average for pairs from the same
cell type minus the average for pairs from different cell types.

@@

The originality of \textit{OneD} lies in the estimation from a single
variable of the explicit biases with a Negative Binomial model and of the
copy number. This allows greater running speed, while preserving a good
performance on samples with karyotypic aberrations. One of the reasons why
\textit{OneD} is able to better highlight the biological distinctions
between samples is that it only corrects the copy number if specifically
requested by the user. The impact of copy number variations on
normalization is rather opaque, especially if they are treated as implicit
biases (Figure~\ref{fig:cnv_correction}B). Treating them as explicit
biases with optional removal seems to be an overall safer strategy. In
homogeneous data sets, correcting for the copy number can blur the
biological signal, but when karyotypes are variable or unstable, it may
increase the reproducibility (Figure~\ref{fig:cnv_correction}A).
If the purpose is to remove the effect of copy number, then OneD+CN
outperforms a karyotype correction method such as caICB because it allows
for more flexible types of aberrations. On the other hand, if the purpose
is to improve the reproducibility among samples of the same type,
correcting the copy number removes part of the signal, and therefore blurs
the distinction between experiments (Figures~\ref{fig:aberrant} and
\ref{fig:normal}).
